---
title: "Western Lake Erie HABs phenology using BHM"
author: "Song Qian"
date: "2026-03-12"
format: pdf
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## Initial setup

```{r}
#| echo: false
#| message: false
#| warning: false
source("FrontMatter_cmdStan.R")
knitr::opts_chunk$set(echo = TRUE)
load(paste(saveDIR, "eriedata_08_21.RData", sep="/"))
load(paste(saveDIR, "eriedata_08_24.RData", sep="/"))
```

## Smoothed Hocky-Stick Model (sHSM)

Modeling a uni-modal seasonal pattern using the smoothed hockey-stick model (sHSM). - sHSM: $$
y = \beta_0+\beta_1(x-x_{mid}) + \delta\cdot scal\cdot \log\left (1+e^{\frac{x-x_{mid}}{scal}}\right )
$$ where $y$ is the log concentration and $x$ is the ordinal date. The model parameter $x_{mid}$ is the date when the peak concentration is reached, $scal$ is the smoothness parameter, $\beta_1$ is the slope (rate of increase on log concentration) before the peak, $\delta$ is the slope difference (i.e., $\beta_1+\delta$ is the slope after peak), and $\beta_0$ is the peak log concentration without smoothing. The smoothed peak is $\beta_0+\delta\cdot scal\cdot \log(2)$. The smoothness parameter ($scal$) and the difference in slope ($\delta$) appear in the equation as a product. As a result, they cannot easily be estimated both as free parameters.

```{r}
#| message: false
packages(VGAM)  ## log1p_exp  ### doesn't work for rv objects
sHSM <- function(x, beta_0, beta_1, beta_2, phi, lambda){
  return(beta_0 + beta_1 * (x - phi) + 
           (beta_2-beta_1) * lambda * 
           log(1+exp((x-phi)/lambda)))
}
```

In its original scale, MC and Chla concentrations are modeled as a function of date ($x$): $$
f(x) = e^{\beta_0+\beta_1(x-x_{mid}) + \delta\cdot scal\cdot \log\left (1+e^{\frac{x-x_{mid}}{scal}}\right )}
$$ The first order derivative is $$
\left(\frac{{\delta}\mathrm{e}^{\frac{x - x_{mid}}{scal}}}{\mathrm{e}^{\frac{x - x_{mid}}{scal}} + 1} + {\beta}_{1}\right) f(x) 
$$ The root of the derivative is $$
scal\cdot \ln\left(-\frac{{\beta}_{1}}{{\delta} + {\beta}_{1}}\right) + x_{mid}
$$ which is the peak location. The second derivative of the model is (checked manually):

$$
\left(\frac{{\delta}\mathrm{e}^{\frac{x - x_{mid}}{scal}}}{scal \left(\mathrm{e}^{\frac{x - x_{mid}}{scal}} + 1\right)} - \frac{{\delta}\mathrm{e}^{\frac{2 \left(x - x_{mid}\right)}{scal}}}{scal \left(\mathrm{e}^{\frac{x - x_{mid}}{scal}} + 1\right)^{2}}\right) f(x) + \left(\frac{{\delta}\mathrm{e}^{\frac{x - x_{mid}}{scal}}}{\mathrm{e}^{\frac{x - x_{mid}}{scal}} + 1} + {\beta}_{1}\right)^{2} f(x)
$$ To find the roots (inflection points), we re-express the second order derivative with respect to $z=\mathrm{e}^{\frac{x-x_{mid}}{scal}}$ as follows: $$
f(x)\left (\frac{\delta z}{scal(1+z)} - \frac{\delta z^2}{scal(1+z)^2} + \left(\beta_1+\frac{\delta z}{1+z} \right )^2\right )
$$ which can be simplified to be a quadratic form with respect to $z$: $$
f(x)\left (\frac{scal(\beta_1+\delta)^2z^2+\left (2scal\beta_1(\beta_1+\delta)+\delta\right )z + scal\beta_1^2}{scal(1+z)^2}\right )
$$ Let $a=scal(\beta_1+\delta)^2, b=2scal\beta_1(\beta_1+\delta) + \delta$, and $c=scal\beta_1^2$, the roots with respect to $z$ are: $$
z_{root} = \frac{-b\pm\sqrt{b^2-4ac}}{2a}
$$ and the inflection points are: $$
x = x_{mid} +scal \log\left( \frac{-b\pm\sqrt{b^2-4ac}}{2a}\right)
$$ To check the accuracy of my algebra, I derived the roots by expressing the second order derivative with respect to $z^{-1}=\mathrm{e}^{-\frac{x-x_{mid}}{scal}}$: $$
f(x)\left (\frac{\delta }{scal(1+z^{-1})} - \frac{\delta}{scal(1+z^{-1})^2} + \left(\beta_1+\frac{\delta}{1+z^{-1}} \right )^2\right )
$$ and the quadratic form with respect to $z^{-1}$ is: $$
f(x)\left (\frac{scal\beta_1^2(z^{-1})^2+(\delta+2scal\beta_1(\beta_1+\delta))z^{-1}+scal(\beta_1+\delta)^2}{scal(1+z^{-1})^2} \right )
$$ The root with respect to $x^{-1}$ is: $$
z_{root}^{-1} = \frac{-b\pm\sqrt{b^2-4ac}}{2c}
$$ To verify the result, I expressed the quadratic form with respect to $z$: $$
az^2+bz+c = a\left (\frac{1}{z^{-1}}\right )^2+b\left (\frac{1}{z^{-1}}\right )+c = \frac{a+bz^{-1}+c(z^{-1})^2}{(z^{-1})^2}
$$ which shows that the derived $z_{root}$ and $z_{root}^{-1}$ are correct.

The inflection point is where the rate of rise (or decline) of MC or Chla concentration is at the maximum. A numerically more efficient way to get the inflection points may be directly calculate the daily first-order derivatives and pick the minimum and the maximum (?).

## hierarchical model for MC/Chla phenology

model: y = b0 + b1(x-Phi)+ (b2)*lambda*Log(1 + e\^(x-phi/lambda))

y = log concentration x = day b0 = y-intercept of peak concentration b1 = rising slope b2 = descending slope (negative) phi = day of peak concentration lambda = smoothness parameter, shape of curve

Compared to the year (temporal) model, the year-site (spatial-temporal) model

```{r}
#| label: BHM_stan
#| eval: false
## with censorship
habs_bhm <- "
data {
  int<lower=0> n; // sample size
  array[n] real x; // day
  int<lower=0> ncens; // censored sample size
  array[ncens] real xcens;  
  int<lower=1> yr;// number of years
  array[n] int<lower=1, upper=yr> year; 
  // year index - each data point assigned to a year
  array[ncens] int<lower=1, upper=yr> yearcens; //
  array[n] real y; // observed log MC/Chla at day x
  array[ncens] real ycens; // detection limits
  array[yr] real lambda; 
  // smoothness parameter, 0.1, different for each year
  vector[yr] xlow; // first day of sampling
  vector[yr] xup; // last day of sampling
}
parameters {
  vector<lower=0>[yr] b1; 
  // rising slope by year (positive)
  vector<upper=0>[yr] b2;
  // descending slope by year (negative)
  vector[yr] b0; // peak concentration
  vector[yr] logit_phi_raw; 
  // raw phi value, fraction of range since beginning - 
  // impose hierarchical structure on this parameter
  // hyperparameters
  real<lower=0> mu_b1; // 
  real<lower=0> sigma_b1;   
  
  real<upper=0> mu_b2; // must be negative                    
  real<lower=0> sigma_b2;    
  
  real mu_b0; // 
  real<lower=0> sigma_b0;   
  
  real mu_phi_raw; //         
  real<lower=0> sigma_phi_raw;
  
  vector<lower=0, upper=10>[yr] sigma; // noise sd indexed by year
}
transformed parameters {
  vector[yr] phi_raw; // phi_raw constrained between 0 - 1
  vector[yr] phi; //
  vector[yr] delta;
  vector[n] mu;
  vector[ncens] mucens;
  vector<lower=0>[n] sig; 
  vector<lower=0>[ncens] sigcens; 
  
  delta = b2 - b1;
  
  for(i in 1:yr){
    phi_raw[i] = inv_logit(logit_phi_raw[i]);
    phi[i] = xlow[i] + (xup[i] - xlow[i]) * phi_raw[i];
    // transformed phi using x low and up
  }
  for (i in 1:n){
    mu[i] = b0[year[i]] + b1[year[i]] * (x[i]-phi[year[i]]) +
      (delta[year[i]]) * lambda[year[i]] *
      log1p_exp((x[i]-phi[year[i]])/lambda[year[i]]); 
    sig[i] = sigma[year[i]]; 
  }
  for (i in 1:ncens){
    mucens[i] = b0[yearcens[i]] + 
      b1[yearcens[i]] * (xcens[i]-phi[yearcens[i]]) + 
      (delta[yearcens[i]]) * lambda[yearcens[i]] * 
        log1p_exp((xcens[i]-phi[yearcens[i]]) / 
        lambda[yearcens[i]]); 
    sigcens[i] = sigma[yearcens[i]]; 
  }
}
model {
  mu_b1 ~ normal(0, 5); // priors for mu
  mu_b2 ~ normal(0, 5);
  mu_b0 ~ normal(0, 5); 
  mu_phi_raw ~ normal(0, 5);
  
  sigma_b1 ~ normal(0, 2); // priors for sigma
  sigma_b2 ~ normal(0, 2);
  sigma_b0 ~ normal(0, 2); 
  sigma_phi_raw ~ normal(0, 2);
  
  sigma ~ normal(0,2);
  
  b1 ~ normal(mu_b1, sigma_b1); 
  b2 ~ normal(mu_b2, sigma_b2);
  b0 ~ normal(mu_b0, sigma_b0); 
  logit_phi_raw ~ normal(mu_phi_raw, sigma_phi_raw);
  
  target += normal_lpdf(y | mu, sig);
  target += normal_lcdf(ycens | mucens, sigcens);
}
"

## without censorship
habs_bhm_nocens <- "
data {
  int<lower=0> n; // sample size
  array[n] real x; // day
  int<lower=1> yr;// number of years
  array[n] int<lower=1, upper=yr> year; 
  // year index - each data point assigned to a year
  array[n] real y; // observed log MC/Chla at day x
  array[yr] real lambda;
  // smoothness parameter, 0.1, different for each year
  vector[yr] xlow; // first day of sampling
  vector[yr] xup; // last day of sampling
}
parameters {
  vector<lower=0>[yr] b1; // rising slope by year (positive)
  vector<upper=0>[yr] b2; 
  // descending slope by year (negative)
  vector[yr] b0; // peak concentration
  vector[yr] logit_phi_raw; 
  // raw phi value, fraction of range since beginning - 
  // impose hierarchical structure on this parameter
  
  // hyperparameters
  real<lower=0> mu_b1; // 
  real<lower=0> sigma_b1;   
  
  real<upper=0> mu_b2; // must be negative                    
  real<lower=0> sigma_b2;    
  
  real mu_b0; // 
  real<lower=0> sigma_b0;   
  
  real mu_phi_raw; //         
  real<lower=0> sigma_phi_raw;
  
  vector<lower=0, upper=10>[yr] sigma; 
  // noise sd indexed by year
}
transformed parameters {
  vector[yr] phi_raw; 
  // phi_raw constrained between 0 - 1
  vector[yr] phi; //
  vector[yr] delta;
  vector[n] mu;
  vector<lower=0>[n] sig; 

  delta = b2 - b1;
  
  for(i in 1:yr){
    phi_raw[i] = inv_logit(logit_phi_raw[i]);
    phi[i] = xlow[i] + (xup[i] - xlow[i]) * phi_raw[i]; 
    // transformed phi using x low and up
  }
  for (i in 1:n){
    mu[i] = b0[year[i]] + b1[year[i]] * (x[i]-phi[year[i]]) +
    (delta[year[i]]) * lambda[year[i]] *
    log1p_exp((x[i]-phi[year[i]])/lambda[year[i]]); 
    sig[i] = sigma[year[i]]; 
  }
}
model {
  mu_b1 ~ normal(0, 5); // priors for mu
  mu_b2 ~ normal(0, 5);
  mu_b0 ~ normal(0, 5); 
  mu_phi_raw ~ normal(0, 5);
  
  sigma_b1 ~ normal(0, 2); // priors for sigma
  sigma_b2 ~ normal(0, 2);
  sigma_b0 ~ normal(0, 2); 
  sigma_phi_raw ~ normal(0, 2);
  
  sigma ~ normal(0,2);
  
  b1 ~ normal(mu_b1, sigma_b1); 
  b2 ~ normal(mu_b2, sigma_b2);
  b0 ~ normal(mu_b0, sigma_b0); 
  logit_phi_raw ~ normal(mu_phi_raw, sigma_phi_raw);
  
  y ~ normal(mu, sig);
}
"
## Compile
habsModel <- write_stan_file(habs_bhm)
habsModel_nocens<-write_stan_file(habs_bhm_nocens)

mod_HABs_cens <- cmdstan_model(habsModel)
mod_HABs_nocens <- cmdstan_model(habsModel_nocens)

save(mod_HABs_cens, mod_HABs_nocens, 
     file=paste(saveDIR,"HabsModelMCChla_cmdStan.RData", sep="/"))
```

Loading compiled models

```{r}
load(paste(saveDIR,"HabsModelMCChla_cmdStan.RData", sep="/"))
# load stan model
```

### Executing the model

1.  Input data function
2.  Running the model

```{r}
#| label: input_function
#| eval: true

infile <- function(data=eriedata08_21, y_col="part_mc",
                   n.chains=nchains,
                   censor=T, censored_at=0.1){
##  data[data[,y_col]<=censored_at, y_col] <- censored_at
  data$yrs <- as.numeric(ordered(format(data$Date,
                                        format="%Y")))
  yr <- max(data$yrs)
  data$days <- as.numeric(format(data$Date, format="%j"))
  xlow <- tapply(data$days, data$yrs, min)
  xup <- tapply(data$days, data$yrs, max)
  
  if(censor){
    temp <- data[,y_col] <= censored_at
    datacens <- data[temp,]
    data2 <- data[!temp,]
    n <- dim(data2)[1]
    ncens=dim(datacens)[1]
    y <- data2[,y_col]
    ycens <- censored_at
    indata <- list(n=n, x=data2$days, y=log(y), 
                   year=data2$yrs, ncens=ncens,
                   xcens=datacens$days,
                   ycens=log(rep(ycens, ncens)),
                   yearcens=datacens$yrs,
                   yr=yr, xlow=xlow, xup=xup, 
                   lambda=0.01*(xup-xlow))
  } else {
    n <- dim(data)[1]
    y <- data[,y_col]
    indata <- list(n=n, x=data$days, y=log(y), year=data$yrs,
                   yr=yr, xlow=xlow, xup=xup, 
                   lambda=0.01*(xup-xlow))
  }
  pars=c("b1", "b2", "phi", "b0", "mu_b0", "mu_b1",
         "mu_b2", "mu_phi_raw", "sigma_b0","sigma_b1",
         "sigma_b2", "sigma_phi_raw", "sigma")
  inits <- list()
  for (i in 1:n.chains){
    tmp = abs(runif(indata$yr))
    tmp2 = abs(runif(1))
    inits[[i]] <- list(
      b1 = tmp,
      b2 = -tmp - abs(runif(indata$yr)),
      b0 = rnorm(indata$yr) + max(log(y))/2,
      logit_phi_raw = rnorm(indata$yr),
      mu_b1 = tmp2,
      mu_b2 = -tmp2 - abs(runif(1)),
      mu_b0 = rnorm(1) + max(log(y))/2,
      mu_phi_raw = rnorm(1),
      sigma_b1 = runif(1),
      sigma_b2 = runif(1),
      sigma_b0 = runif(1),
      sigma_phi_raw = runif(1),
      sigma = runif(indata$yr)         
    )
  }
  return(list(data=indata, inits=inits, pars=pars))
}
```

```{r}
#| label: Running the model
#| eval: false
input.to.stan <- infile(n.chains = nchains)
### run model ### about 460/990 seconds on ThinkStation/miniPC
fitall <- mod_HABs_cens$sample( 
                 data = input.to.stan$data, 
                 init = input.to.stan$inits,
                 parallel_chains = nchains,
                 chains = nchains,
                 iter_warmup = niters/2,
                 iter_sampling = niters/2,
                 thin = nthin, 
                 refresh = niters/10, 
                 output_dir = saveDIR)
fitall$summary(variables=input.to.stan$pars)
fitall$summary(variables=c("mu_b0","mu_b1","mu_b2","mu_phi_raw"))

input.to.stan24 <- infile(data=eriedata08_24, n.chains = nchains)
### run model ### about 700/1200 seconds on ThinkStation/miniPC
fitall24<-mod_HABs_cens$sample( 
                 data = input.to.stan24$data, 
                 init = input.to.stan24$inits,
                 parallel_chains = nchains,
                 chains = nchains,
                 iter_warmup = niters/2,
                 iter_sampling = niters/2,
                 thin = nthin, 
                 refresh = niters/10, 
                 output_dir = saveDIR)

fitall24$summary(variables=input.to.stan24$pars)
fitall24$summary(variables=c("mu_b0","mu_b1","mu_b2","mu_phi_raw"))

### run model ### about 700/1840 seconds on ThinkStation/miniPC
chla_0 <- is.na(eriedata08_24$Chla) | eriedata08_24$Chla<=0.01
input.to.stan.chla <- infile(data=eriedata08_24[!chla_0, ],
                             y_col="Chla", 
                             censor = F, n.chains = nchains)

fitall_chla <- mod_HABs_nocens$sample( 
                 data = input.to.stan.chla$data, 
                 init = input.to.stan.chla$inits,
                 parallel_chains = nchains,
                 chains = nchains,
                 iter_warmup = niters/2,
                 iter_sampling = niters/2,
                 thin = nthin, 
                 refresh = niters/10, 
                 output_dir = saveDIR)

fitall_chla$summary(variables=input.to.stan.chla$pars)
fitall_chla$summary(variables=c("mu_b0","mu_b1","mu_b2","mu_phi_raw"))


save(fitall, fitall24, fitall_chla, 
     file=paste(saveDIR, "fitall1_cmdstan.RData", sep="/"))
```

```{r}
#| label: model_results
input.to.stan <- infile()
input.to.stan24 <- infile(data=eriedata08_24)
chla_0 <- is.na(eriedata08_24$Chla) | eriedata08_24$Chla<=0.01
input.to.stan.chla <- infile(data=eriedata08_24[!chla_0, ],
                             y_col="Chla", 
                             censor = F, n.chains = nchains)
save(input.to.stan, input.to.stan24, chla_0, input.to.stan.chla, file=paste(saveDIR, "input_to_stan.RData", sep="/"))
```

